RVD2

### 2013-08-28 doc/bioinformatics2013/RVD2.tex
paper draft with update on statistic inference.

\documentclass[11pt,reqno]{amsart}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{algorithm, algorithmic}
\usepackage{graphicx}

\usepackage{amssymb}
\usepackage{amsmath}
%\usepackage{epstopdf}
%\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

@@

For each genomic location,  $j=1,\ldots,J$, and each repilcate  $i=1,\ldots,N$, $n_{ij}$ is the total counts, while $r_{ij}$ stands for the non-reference read counts. Variable $r_{ij}$ is drawn from a binomial distribution with probability parameter $\theta_{ij}$ and total count $n_{ij}$.  The error probability $\theta_{ij}$ has a prior beta distribution with position -specific rate parameter $mu_j$ and precision parameter $M_j$.

The position error rate $\mu_j$ has a beta distribution as hypoprior distribution with parameter $\mu_0$ and $M_0$. This is to ensure that the error rate is between 0 and 1. The prior is useful for situations when there is a significant minor allele.

@@

Instead of attempting to directly compute the posterior distribution, we instead sample it using a Metropolis-Hastings within Gibbs scheme as discribed in Metropolis-within-Gibbs section.

@@

The estimate of parameter $\mu_j$ is obtained using the expected value of $\theta_{ij}$, while $M_j$ is estimated using the moment esitmates for the variance of $\theta_{ij}$.
\begin{equation}
 \mu_j=E(\theta_{ij}|\mu_j,M_j)
\end{equation}
\begin{equation}
 M_j=\frac{\mu_j(1-\mu_j)}{Var(\theta_{ij}|\mu_j,M_j)}-1
\end{equation}

@@

Because of the computation complexition of the nomalization factor in the posterior probability, A Markov Chain Monte Carlo (MCMC) sampling approch is employed to approximate the posterior distribution $p(\theta, \mu | r,n; \mu_0,M_0,M)$. Specifically Metropo-within-Gibbs algorithm is used to obtain the posterior distribution of each latent variable.

\begin{algorithm}[ht]
\caption{Metropolis within Gibbs Algorithm}
\label{alg:metro_gibbs}
\begin{algorithmic}[1]
\REPEAT
\FOR {j = 1 to J}
  \FOR {i =1 to N}
	\STATE Draw 1 samples from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$ using Gibbs sampling
  \ENDFOR
\ENDFOR

\FOR {j = 1 to J}
	\STATE Draw T samples from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$ using M--H
	\STATE Set $mu_j$ to the sample median
\ENDFOR
\UNTIL {sample size is sufficient}

@@

\subsection{Bayesian Hypothesis Testing}

>>>>

\documentclass[11pt,reqno]{amsart}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
%\usepackage{algorithm, algorithmic}

\usepackage{algorithm}
\usepackage{algpseudocode}

\usepackage{graphicx}

\usepackage{verbatim}
\usepackage{amssymb}
\usepackage{amsmath}
%\usepackage{epstopdf}
%\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

@@

For each genomic location,  $j=1,\ldots,J$, and each replicate  $i=1,\ldots,N$, $n_{ij}$ is the total counts, while $r_{ij}$ stands for the non-reference read counts. Variable $r_{ij}$ is drawn from a binomial distribution with probability parameter $\theta_{ij}$ and total count $n_{ij}$.  The error probability $\theta_{ij}$ has a prior beta distribution with position -specific rate parameter $mu_j$ and precision parameter $M_j$.

The average positional error rate $\mu_j$ across replicates has a beta distribution as hyperprior distribution with parameter $\mu_0$ and $M_0$. This is to ensure that the error rate is between 0 and 1. The prior is useful for situations when there is a significant minor allele.

@@

Instead of attempting to directly compute the posterior distribution, we instead sample it using a Metropolis-Hastings within Gibbs scheme as described in Metropolis-within-Gibbs section.

@@

The estimate of parameter $\mu_j$ is obtained using the expected value of $\theta_{ij}$, while $M_j$ is estimated using the moment estimates for the variance of $\theta_{ij}$.
\begin{equation}
 \mu_j=E(\theta_{ij}|\mu_j,M_j)
\end{equation}
\begin{equation}
 M_j=\frac{\mu_j(1-\mu_j)}{Var(\theta_{ij}|\mu_j,M_j)}-1
\end{equation}

@@

Because of the computation complexities of the normalization factor in the posterior probability, A Markov Chain Monte Carlo (MCMC) sampling approach is employed to approximate the posterior distribution $p(\theta, \mu | r,n; \mu_0,M_0,M)$. Specifically Metropolis-within-Gibbs algorithm is used to obtain the posterior distribution of each latent variable.

\begin{algorithm}[ht]
\caption{Metropolis within Gibbs Algorithm}
\label{alg:metro_gibbs}
\begin{algorithmic}[1]
\Repeat
\For {j = 1 to J}
  \For {i =1 to N}
	\State Draw 1 samples from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$ using Gibbs; sampling
  \EndFor
\EndFor

\For {j = 1 to J}
	\State Draw T samples from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$ using M--H;
	\State Set $mu_j$ to the sample median;
\EndFor
\Until {sample size is sufficient}

@@

\subsection{Statistic Inference}\

Full Bayesian Hypothesis Testing is done to test rare variants in the model.

\subsubsection{Posterior distribution difference sampling}\

Using MCMC sampling, we can obtain the posterior distribution $p \left( \mu |\theta,\mu_0,M_0\right)$ for reference sample and testing sample. In order to compare the difference of position error rate $\mu$ between reference sample and testing sample , a random sampling of posterior distribution is done 
\begin{algorithm}
\caption{Posterior distribution difference sampling}
\begin{algorithmic}[1]
\Procedure{}{}
\Repeat
\State  Randomly  draw one Gibbs sample from $p \left( \mu |r, n, M, \mu_0, M_0\right)$ from reference samples
\State  Randomly draw one Gibbs sample from $p \left( \mu |r, n, M, \mu_0,M_0\right) $ from testing samples
\State Subtract the former from the later to get one sample for the posterior difference distribution 
\Until {Sample size is sufficient}
\EndProcedure
\end{algorithmic}
\end{algorithm}

With the posterior difference distribution of position error rate $\mu$ across replicates obtained, Bayesian inference can be done to summarize the difference.

$p \left( \mu_{testing}-\mu_{ref} |r,n; \mu_0,M_0\right)$

\subsubsection{Bayesian Hypothesis Testing}
To detect a mutation in a specific position, a threshold $\tau$ is set for the posterior difference distribution testing.
\begin{align}
 H_1: \mu_{testing}-\mu_{ref}\geq\tau \\
 H_2: \mu_{testing}-\mu_{ref}<\tau
\end{align}

We can calculate the posterior probability of the hypothesis $H_1$ being true by integrating the posterior density over the correct region:

\begin{equation}
 p(H1:\mu_{testing}-\mu_{ref}\geq\tau)=\int_\tau^\infty p(\mu_{diff} |r,n; \mu_0,M_0)d\mu_{diff}
\end{equation}

We accept the hypothesis $H_1$ if this posterior probability is no lower than 95\%, which means we are 95\% sure that there is a mutation in this position.

ROC curve is plotted to evaluate the performance of the detection algorithm with different threshold for different mutation rate.

\subsubsection{Uniformity Testing}
